Software Development in Startup Companies: The Green-field Startup Model

### 2013-03-04 article/main.tex
Compiles again...

To maintain the development activities, oriented to limited but suitable 
functionality, studies suggest to externalize the complexity of parts of the 
project to third party solutions by means of outsourcing activities 
\cite{Hanna2010}, software reuse \cite{Jansen2008} and open-source strategies , 
\cite{Wall2001, Bean2005}. 
%MUN: Can you split up the references? I.e. write:
%"outsourcing activities \cite{Bla}, software reuse \cite{Bla}, and open-source 
%strategies \cite{Bla}.
%CG: fixed
%------------------------------------------------------------------------------%
\subsection{Technical debt}
Notwithstanding, a new area of the SE research, trying to tackle the problem of 
the \textit{technical debt},  brings and encompasses different implications in 
studying development in software startups \cite{Tom2013}. The metaphoric 
neologism was originally introduced by Cunningham in 1992 
\cite{TechnicalDebtCunn} and has recently attracted the attention of SE 
researchers\footnote{To attest the fact that \textit{technical debt} is gaining 
traction among researchers, we mention two important contributions which 
characterize the \textit{``debt landscape''} : \cite{Nugroho2011,Izurieta2012} 
and a dedicated workshop \cite{workshopDebt} organized by the Software 
Engineering Institute  and ICSE.}. The concept of \textit{technical debt} is 
well illustrated in \cite{Brown:2010:MTD:1882362.1882373} : \textit{``The idea 
is that developers sometimes accept compromises in a system in one dimension 
(e.g., modularity) to meet an urgent demand in some other dimension (e.g., a 
deadline), and that such compromises incur a ``debt'' on which ``interest'' has 
to be paid and which the ``principal'' should be repaid at some point for the 
long-term health of the project''}.  The compromise between high-speed and 
high-quality engineering is faced daily by software startups, not only in terms 
of architecture design but in multifaceted aspects (weak project management, 
testing, process control).

@@

\begin{compactitem}
\item Software development strategy: the overall approach adopted by the 
company to carry out the product development.
\item Engineering activities: the activities needed to bring a product from 
idea to market. Traditional engineering activities are, among others, 
requirement engineering, design, architecture, implementation, testing.
\item Engineering elements: any practice, tool and artifacts contributing and 
supporting the engineering activities.
\item Quality attributes: those overall factors that affect run-time behavior, 
system design, and user experience. They represent areas of concern that have 
the potential for applications to impact across layers and tiers. Some of these 
attributes are related to the overall system design, while others are specific 
to run time, design time, or user centric issues \cite{Microsoft2009}.
\item Growth: an increase in the company size respect to the initial conditions 
in terms of either employees or users/customers, and product complexity in terms 
of handling an increasing number of feature requests.
\item Software product: any software product and software service.
\end{compactitem}
%------------------------------------------------------------------------------%
\section{Research methodology}
\label{resmet}

\begin{figure*}[!t]
\centering
\includegraphics[width=5in]{figures/completemethodology2}
\caption{Grounded theory study process overview}
\label{fig:gt:completemethodology}
\end{figure*}
%MUN: I think you can increase the font size in the figure a bit. Try thereby
%to keep the height of the figure, expanding however the width (use horizontal
%space since it is a two column figure). Change also in figure:  - Write "Data
%Analysis" horizontally  - Change names of macro-phases to upper-case, i.e.
%Design and Execution,     Data Collection, Data Analysis, etc.  - Add section
%number to macro-phase name, i.e. 3.1 Design and Execution, 3.2     Data
%Collection  - Change "Discussion and Validation" into "Theory Validation" as
%you name it     in the text. Also, add a section (3.5) that describes shortly
%what you did in "Theory  Validation" 
%CG: fixed... fontsize cannot be increased uniformly among all the boxes, then 
%I skipped that part...
%MUN1: OK, but in the figure, the correct section number for theory validation 
%is 3.5, not 3.4

@@

\item RQ-1 : What is the current state-of-practice related to software 
development strategy in early-stage startups?
\begin{compactitem}

\item RQ-1.1 : How do startups structure and execute the main engineering 
activities?
\item RQ-1.2 : How are product quality attributes considered by startups?
\end{compactitem}

@@

The follow-up questionnaires were designed to capture additional data, gather 
missing information and confirm interview results by triangulation. The 
questionnaires have been tailored to each startup, partially taking advantage of 
the repertory grid principles \cite{Edwards2009}. To obtain better chances of 
quickly obtaining responses, questionnaires have been sent to respondents 
immediately after transcribing the interview results. Most of the follow-up  
questionnaires have been collected within one day after the conclusion of the 
last interview.   
%MUN: I don't understand this: you
%collected the data from questionnaires after ALL interviews have been 
%finished, but on the other hand you said the overall process was iterative 
%(data analysis and triangulation let themes emerge which in turn were used
%in subsequent interviews). 
%CG: fixed...

@@

\section{Results: Green-field startup model}
\label{res:gsm}

@@

The GSM captures the underlying phenomenon of software development in 
early-stage startups. The framework is formed by 128 sub-categories, clustered 
in 35 groups, and finally in 7 categories at the highest level of abstraction. 
By the means of the GSM we provide explanations of the development strategies 
and engineering activities undertaken by startups.
%------------------------------------------------------------------------------%
\subsection{Framework overview}
\label{res:gsm:frmov}
The main concepts representing the underlying phenomenon have been grouped 
together to form high-level categories. Figure \ref{fig:gsm} shows the network 
of causal relationships (represented by arrows) between categories (represented 
by blocks). Each node is linked to subsequent nodes which are called successors, 
and preceding nodes, called predecessors. The relations between them are denoted 
by the direction of the interconnecting arrows. The network is read from left to 
right.

@@

In forthcoming the explanation of the GSM we emphasize labels of theoretical 
categories using the \textit{italic} font style. For the main categories shown 
in Figure \ref{fig:gsm}, when necessary we also make use of identifiers (i.e. 
CATx) to facilitate the reading of this section.
The network is centered around the core category, \textit{speed-up 
development}, which is the most interconnected node in the framework reflecting 
the fact that \textit{``it is the one (category) with the greatest explanatory 
power''}  \cite{Strauss1998}. The meaning of the labels on the three arrows that 
reach the core category (efficiency, effectiveness and performance) is discussed 
in detail in section \ref{res:val}. %MUN: this refers to the validity threats 
%section. Do you mean rather the "Theory Validation" section?
%CG:fixed
%MUN1: this should be \ref{sect:theory:validation:stats}, right? Section 5 is 
%quite long and you should point to the correct subsection

@@

The \textit{severe lack of resources} forces the company to focus on 
implementing an essential set of functionalities and it is one of the main 
reasons why the \textit{product quality has low priority} respect to other more 
urgent needs\footnote{There are some exceptions where the quality aspects 
actually matter and such cases will be discussed in section \ref{res:val}%MUN: 
%this refers to the validity threats section. Do you mean rather the "Theory 
%Validation" section?
%CG fixed
%MUN1: Same here, point to the correct subsection.
.}. At the same time, to be able to deal with such constraints, startups must 
depend on a small group of capable and motivated individuals.

@@

The initial employees are the ingredients which enable high levels of 
performance in software development. To support a fast-paced production 
environment, engineers are required to be highly committed, co-located, 
multi-role, and self-organized. In other words \textit{team is the catalyst of 
development}.
With an essential and flexible workflow, which relies on tacit knowledge 
instead of formal documentation, startups are able to achieve very short 
time-to-market cycles. However, each line of code, written without following 
structures and processes, contributes to grow the \textit{accumulated technical 
debt}, which is further increased by having almost non-existing specifications, 
a minimal project management and a lack of automated tests.

@@

%------------------------------------------------------------------------------%
\subsection{Severe lack of resources}
\label{res:gsm:cat7}
The concept of \textit{severe lack of resources} characterizes the uncertainty 
of the development strategies in startups and it is composed by three 
subcategories: \textit{time-shortage}, \textit{limited human resources} and 
\textit{limited access to expertise}.

@@

Early engineers in the founding team of a startups are typically 
\textit{multi-role and full-stack engineers}: they are full-stack for being able 
to tackle different problems at different levels of the technology stack 
(\textit{generalists developers instead of specialists}) and multi-role since 
usually \textit{engineers are responsible for marketing/sales/development}. As 
remarked by  C11: \textit{``Instead of hiring gurus in one technology, startups 
should hire young developers, generalists, that know how to quickly learn new 
technologies, and quickly move among a huge variety of tasks.''}

@@

A \textit{limited access to expertise} forces the team to rely mainly on their 
personal abilities, even though asking opinions to mentors is reported to be a 
viable practice to maintain feasible objectives. Furthermore \textit{teams work 
under constant pressure}, \textit{``up to 16 hours per day''}, %MUN: add Cx to 
%this statement since it is a citation? 
%CG: not fixed. It represents a subcategory, expressed by the 70% of the 
%interviewed startups
%MUN1: and how should the reader know that? See my problem with italics: you 
%use italics for both interview statements and (sub-)categories. 
mainly constrained by a tight \textit{time shortage}.

Finally, startups present founders-centric structures, and especially in the 
early-stage, it is clear that \textit{CTO/CEO background has high-impact} on the 
company's development approach. For instance, in case of an academic background, 
the CTO encourages the introduction of some architectural design before the 
development phase. Moreover, even though the CTO/CEO starts guiding initially 
the development process, most of the decision are taken consensually by all 
members of the team. Then, the CTO/CEO decides only in particular situations 
where some conflicts occur.
%------------------------------------------------------------------------------%
\subsection{Evolutionary approach}
\label{res:gsm:cat2}
Among different approaches to development, startups prefer to build an initial 
prototype and iteratively refine it over time, similarly to the concept of 
\textit{``evolutionary prototyping''} \cite{EvProt}. The reason of using this 
approach is that startups want to validate the product against the market as 
soon as possible, finding the proper product/market fit. Indeed, they can focus 
on developing only parts of the system they want to validate instead of working 
on developing a whole new system. Then, as the prototype is released, users 
detect opportunities for new functionalities and improvements, providing their 
feedback to developers.

@@

The objective of this evolutionary approach is to avoid wasting time in 
``over-engineering the system'' and building complex functionalities that have 
not been tested on real users. By releasing a very small number of good-enough 
functionalities (see CAT3) the startup can verify the suitability of the 
features and understand how to adjust the direction of product development 
towards actual users' needs. The first version of the product is typically a 
prototype which contains the basic functionalities developed with the least 
possible effort that can validate critical features, enabling the startup's 
survival in the short term (recalling  Ries' definition of \textit{Minimum 
Viable Product} (MVP) \cite{Ries2011}).

@@

%------------------------------------------------------------------------------%
\subsection{Product quality has low priority}
\label{res:gsm:cat3}
The interests of software startups, related to the product, are mainly 
concentrated on building a  \textit{limited number of suitable functionalities} 
rather than fulfilling restrictive non-functional requirements. This strategy 
allows them to quickly release simple products with less need of preliminary 
architectural studies.

@@

Finally, startups implement a \textit{limited number of suitable 
functionalities}, which in view of \textit{limited human resources} and 
\textit{time shortage}, represent a viable strategy to focus on shortening 
time-to-market.
%------------------------------------------------------------------------------%
\subsection{Speed-up development}
\label{res:gsm:cat1}
As remarked initially, \textit{speed-up development} represents the core 
category of the GSM. Firmly grounded as the primary objective of startups, it 
shows the most important characteristic of developing software in the early 
stage.

@@

Finally the constant pressure under which the company regularly operates, leads 
the team to often \textit{work overtime to meet deadlines}. But as reported by 
practitioners, such way of working can be an effective strategy only in the 
short term since it can lead to poorly maintainable code and developer burnouts 
in the long run.
%------------------------------------------------------------------------------%
\subsection{Accumulated technical debt}
\label{res:gsm:cat5}
Startups achieve a very high development speed by radically ignoring aspects 
related to documentation, structures and processes. However, although these 
aspects are not considered important in the very first stages, they will become 
increasingly more important for the productivity in the long-term, as we 
illustrate in CAT6.

@@

Another important factor which contributes to \textit{accumulate the technical 
debt} is the general \textit{lack of architectural design}, substituted by 
\textit{high-level mockups and low-precision diagrams}, \textit{using 
modularization and frameworks from day 1} and \textit{describing critical 
interactions with third-party components only}. In particular, the use of 
well-known standards, frameworks and conventions limits the need of formal UML 
\cite{UML} diagrams and documentation, and provide a minimum level of 
maintainability costs. Additionally, since quality aspects are not a main 
concern (see \textit{limited number of functionalities} and \textit{efficiency 
emerges after using the product} in CAT3) having a well-structured architecture 
remains a secondary priority, and design is conducted only when strictly 
necessary.

@@

Furthermore, a rigid project management is perceived as \textit{``waste of 
time''} which hinders the development speed since the \textit{uncertainty makes 
formal scheduling pointless} (C9  reports that \textit{``initial chaos helps to 
develop faster''}). Startups' \textit{minimal project management}  is supported 
by keeping: \textit{internal milestones short and informal}, 
\textit{low-precision task assignment mechanism}  and a very low attention for 
project metrics (paraphrasing C13, \textit{``the only track of progress was made 
by looking at closed tickets''}). In this context \textit{only a final release 
milestone is viable}, which helps practitioners to remain focused on short term 
goals and put new features in production. To further \textit{speed-up 
development}, startups \textit{simplify issue management by integrating it with 
feature management}. Then, the \textit{limited number of functionalities} and 
the {use of standard/known technologies} with a \textit{simple workflow} are the 
main reasons for 
not establishing a heavy project management process since the project and 
technologies can be autonomously managed by the development team (with 
characteristics of: co-location, self-organization and very small team size). 
%MUN: do you mean small team size?
%CG: fixed

Finally, one of the categories, which most contributes in growing the 
\textit{accumulated technical debt}, is the substantial  use of informal and 
verbal communication channels on a daily basis. The high co-location and the 
fast paced development approach increase the volume of \textit{tacit knowledge} 
and severe lack of any kind of documentation.  C4 observes in this regard that:  
\textit{``[\ldots] the issue of having documentation and diagrams out of the 
source code is that you need to update them every time you change something. 
There is no time for it. Instead, there is a huge pay off in having a code that 
is understandable itself.''}
This approach is supported by the fact that startups make use of  
\textit{simple and informal workflow}, \textit{standard/known technologies}, 
\textit{very small and co-located development team} with \textit{limited need of 
formalities}.
%------------------------------------------------------------------------------%
\subsection{Initial growth hinders performance}
\label{res:gsm:cat6}
The lack of attention given in the first phases to engineering activities 
allows startups to ship code extremely quickly (see CAT5). However, if 
successful, the initial product becomes more and more complex over time, the 
number of users increases and the company size starts to grow. Under these 
circumstances the need of controlling the initial chaos forces the development 
team to return the \textit{accumulated technical debt} instead of focusing on 
new users' requests. Hence, \textit{initial growth hinders performance} in terms 
of number of new user stories implemented in a certain amount of time, which 
brings new suitable functionalities to the users.

@@

Therefore, startups begin to \textit{partially replace informal communication 
with traceable systems} and \textit{introduce basic metrics for measuring 
project and team progress} in order to establish an initial structured workflow. 
In conclusion, because the increasing number of users causes issues in 
scalability and reliability, the \textit{need of more structured workflow} 
becomes of major relevance. C11 states: \textit{``Yet, it is still better to 
have a reasonable drop-down in performance when team grows than lose time in the 
beginning''}.
%------------------------------------------------------------------------------%
\subsection{Theory}
\label{res:gsm:th}
In order to explain and understand the development strategies in early-stage 
software startups we construct the theory generated and supported by the above 
presented GSM:

@@

In this section we discuss the validity of the GSM by means of 
cross-methodological observations, as discussed in section \ref{rm:val}.
%MUN: expand this discussion a bit (motivation why you do this like you do it) 
%and move to Section 3.5. Then write here just that you perform theory 
%validation as discussed in Section 3.5 (and remove the bullet list below).
%CG:fixed
%MUN1: Ok, but fix section 3.5... you see that you don't motivate 5.1-5.4 at 
%all there and that the current content of 3.5 is not related to what you 
%actually do here as validation?

@@

Starting from the Coleman's work, we refer to him since only he has,  to the 
best of our knowledge, conducted a similar research in startups' context, even 
though with a different focus. Coleman investigated factors in software 
development that hinder initiatives of software process improvement (SPI) in a 
later stage, representing also companies in expansion phase with more than 100 
employees. %MUN: so he did not do research 
%in a startup context? You need to clarify what you mean by "more mature 
%companies"... I assume you still mean startups, but startups at a later stage, 
%i.e. after first product release.
%CG: fixed

@@

As also reported in the GSM, the definition of a \textit{``minimum process''} 
is not a matter of poor knowledge and training, but rather it is the necessity 
of operating with solutions that let the company move faster. 
\textit{``One-size-fits-all''} solutions have always found difficulty in 
penetrating small software organizations \cite{Staples2007}. When startups began 
establishing any SPI process, they experienced \textit{process erosions}  
\cite{Coleman2008} %MUN: 
%add ref (Coleman?). You should add, for each category you use from Coleman, 
%add the reference to the paper. This makes it clear that the category comes 
%from him (and is not one of our categories)
%CG: fixed
, which resolved to a barely sufficient workflow to satisfy the organizational 
business needs.

Then, software startups favor the use of agile principles in support of 
\textit{creativity} and \textit{flexibility} instead of SPI methods, whereby 
processes need to be predictable and repeatable. Nimble and ad-hoc solutions 
prevent the use of heavy bureaucracy and formal communication strategies %MUN: 
%seems to contradict each other: "ad-hoc solutions prevent the use of heavy 
%bureaucracy AND fast communication strategies" I agree that ad-hoc solutions 
%do not produce heavy bureaucracy, but they allow fast communication 
%strategies... I may be wrong, but then you need to explain it better
%CG:fixed, you are right :) ...  
, even though the accumulated tacit knowledge is hard to manage and to transfer 
to new hires as also discussed in category \textit{CAT6}.

@@

\subsubsection{Market requirements and application type} 
Market requirements \cite{Coleman2007} and application type \cite{Sutton2000, 
Coleman2008, Coleman2008a} are two further potential confounding factors to the 
GSM. Their main impact is related to the adoption of flexible and reactive 
solutions for the development process. In particular, they refer to the 
necessity of fulfillment of quality concerns that goes beyond scalability and UX 
defined by the GSM. Especially when requirements are rigidly imposed or the 
application domain is well-known, providing low-quality products to final users 
might determine the failure of a startup.

>>>>

To maintain the development activities, oriented to limited but suitable
functionality, studies suggest to externalize the complexity of parts of the
project to third party solutions by means of outsourcing activities
\cite{Hanna2010}, software reuse \cite{Jansen2008} and open-source strategies ,
\cite{Wall2001, Bean2005}.  %MUN: Can you split up the references? I.e. write:
%"outsourcing activities \cite{Bla}, software reuse \cite{Bla}, and open-source
%strategies \cite{Bla}. %CG: fixed

%------------------------------------------------------------------------------%
\subsection{Technical debt} Notwithstanding, a new area of SE research,
trying to tackle the problem of the \textit{technical debt}, brings and
encompasses different implications in studying development in software startups
\cite{Tom2013}. The metaphoric neologism was originally introduced by
Cunningham in 1992 \cite{TechnicalDebtCunn} and has recently attracted the
attention of SE researchers\footnote{To attest the fact that \textit{technical
debt} is gaining  traction among researchers, we mention two important
contributions which characterize the \textit{``debt landscape''}:
\cite{Nugroho2011,Izurieta2012} and a dedicated workshop \cite{workshopDebt}
organized by the Software  Engineering Institute  and ICSE.}. The concept of
\textit{technical debt} is  well illustrated in
\cite{Brown:2010:MTD:1882362.1882373}: \textit{``The idea  is that developers
sometimes accept compromises in a system in one dimension  (e.g., modularity) to
meet an urgent demand in some other dimension (e.g., a  deadline), and that such
compromises incur a ``debt'' on which ``interest'' has  to be paid and which the
``principal'' should be repaid at some point for the long-term health of the
project''}.  The compromise between high-speed and high-quality engineering is
faced daily by software startups, not only in terms  of architecture design but
in multifaceted aspects (weak project management, testing, process control).

@@

\begin{compactitem} 
\item Software development strategy: the overall approach adopted by the  
company to carry out the product development. 
\item Engineering activities: the activities needed to bring a product from  
idea to market. Traditional engineering activities are, among others,  
requirement engineering, design, architecture, implementation, testing. 
\item Engineering elements: any practice, tool and artifacts contributing and  
supporting the engineering activities.
\item Quality attributes: those overall factors that affect run-time
behavior, system design, and user experience. They represent areas of concern
that have  the potential for applications to impact across layers and tiers.
Some of these  attributes are related to the overall system design, while others
are specific  to run time, design time, or user centric issues
\cite{Microsoft2009}. 
\item Growth: an increase in the company size respect to the initial 
conditions in terms of either employees or users/customers, and
product complexity in terms  of handling an increasing number of feature
requests. 
\item Software product: any software product and software service.
\end{compactitem}

%------------------------------------------------------------------------------%
\section{Research methodology} \label{resmet}

\begin{figure*}[!t] \centering
\includegraphics[width=5in]{figures/completemethodology2} \caption{Grounded
theory study process overview} \label{fig:gt:completemethodology} \end{figure*}
%MUN: I think you can increase the font size in the figure a bit. Try thereby
%to keep the height of the figure, expanding however the width (use horizontal
%space since it is a two column figure). Change also in figure:  - Write "Data
%Analysis" horizontally  - Change names of macro-phases to upper-case, i.e.
%Design and Execution,     Data Collection, Data Analysis, etc.  - Add section
%number to macro-phase name, i.e. 3.1 Design and Execution, 3.2     Data
%Collection  - Change "Discussion and Validation" into "Theory Validation" as
%you name it     in the text. Also, add a section (3.5) that describes shortly
%what you did in "Theory  Validation"  %CG: fixed... fontsize cannot be
%increased uniformly among all the boxes, then  I skipped that part... 
%MUN1: OK, but in the figure, the correct section number for theory validation  
%is 3.5, not 3.4. CG:fixed

@@

\item RQ-1 : What is the current state-of-practice related to software
development strategy in early-stage startups?

  \begin{compactitem}
  \item RQ-1.1 : How do startups structure and execute the main engineering
    activities? 
  \item RQ-1.2 : How are product quality attributes considered by
    startups? 
  \end{compactitem}

@@

The follow-up questionnaires were designed to capture additional data, gather
missing information and confirm interview results by triangulation. The
questionnaires have been tailored to each startup, partially taking advantage of
the repertory grid principles \cite{Edwards2009}. To obtain better chances of
quickly obtaining responses, questionnaires have been sent to respondents
immediately after transcribing the interview results. Most of the follow-up
questionnaires have been collected within one day after the conclusion of the
last interview.    
%MUN: I don't understand this: you %collected the data from questionnaires 
%after ALL interviews have been finished, but on the other hand
%you said the overall process was iterative  (data analysis and triangulation
%let themes emerge which in turn were used %in subsequent interviews).  
%CG: fixed...

@@

\section{Results: Green-field startup model}\label{res:gsm}

@@

The GSM captures the underlying phenomenon of software development in  early-
stage startups. The framework is formed by 128 sub-categories, clustered  in 35
groups, and finally in 7 categories at the highest level of abstraction.  By the
means of the GSM we provide explanations of the development strategies  and
engineering activities undertaken by startups.
%------------------------------------------------------------------------------%
\subsection{Framework overview} \label{res:gsm:frmov} The main concepts
representing the underlying phenomenon have been grouped  together to form high-
level categories. Figure \ref{fig:gsm} shows the network  of causal
relationships (represented by arrows) between categories (represented  by
blocks). Each node is linked to subsequent nodes which are called successors,
and preceding nodes, called predecessors. The relations between them are denoted
by the direction of the interconnecting arrows. The network is read from left to
right.

@@

In forthcoming the explanation of the GSM we emphasize labels of theoretical
categories using the \textit{italic} font style. For the main categories shown
in Figure \ref{fig:gsm}, when necessary we also make use of identifiers (i.e.
CATx) to facilitate the reading of this section. The network is centered around
the core category, \textit{speed-up  development}, which is the most
interconnected node in the framework reflecting  the fact that \textit{``it is
the one (category) with the greatest explanatory  power''}  \cite{Strauss1998}.
The meaning of the labels on the three arrows that  reach the core category
(efficiency, effectiveness and performance) is discussed  in detail in section
\ref{sect:theory:validation:stats}. %MUN: this refers to the validity threats
%section. Do you mean rather the "Theory Validation" section? %CG:fixed 
%MUN1: this should be \ref{sect:theory:validation:stats}, right? Section 5 is  
%quite long and you should point to the correct subsection. CG: fixed.

@@

The \textit{severe lack of resources} forces the company to focus on
implementing an essential set of functionalities and it is one of the main
reasons why the \textit{product quality has low priority} respect to other more
urgent needs\footnote{There are some exceptions where the quality aspects
actually matter and such cases will be discussed in section \ref{sect:an:comp
:cat-vs-literature:conf}
%MUN: this refers to the validity threats section. Do you mean rather the 
%"Theory Validation" section? %CG fixed 
%MUN1: Same here, point to the correct subsection. CG:fixed 
.}. At the same time, to be able to deal with such constraints, startups must  
depend on a small group of capable and motivated individuals.

@@

The initial employees are the ingredients which enable high levels of
performance in software development. To support a fast-paced production
environment, engineers are required to be highly committed, co-located,  multi-
role, and self-organized. In other words \textit{team is the catalyst of
development}. With an essential and flexible workflow, which relies on tacit
knowledge  instead of formal documentation, startups are able to achieve very
short  time-to-market cycles. However, each line of code, written without
following  structures and processes, contributes to grow the \textit{accumulated
technical  debt}, which is further increased by having almost non-existing
specifications,  a minimal project management and a lack of automated tests.

@@

%------------------------------------------------------------------------------%
\subsection{Severe lack of resources}\label{res:gsm:cat7} 
The concept of \textit{severe lack of resources} characterizes the uncertainty  
of the development strategies in startups and it is composed by 
three subcategories: \textit{time-shortage}, \textit{limited human resources} 
and \textit{limited access to expertise}.

@@

Early engineers in the founding team of a startups are typically  \textit{multi-
role and full-stack engineers}: they are full-stack for being able  to tackle
different problems at different levels of the technology stack
(\textit{generalists developers instead of specialists}) and multi-role since
usually \textit{engineers are responsible for marketing/sales/development}. As
remarked by  C11: \textit{``Instead of hiring gurus in one technology, startups
should hire young developers, generalists, that know how to quickly learn new
technologies, and quickly move among a huge variety of tasks.''}

@@

A \textit{limited access to expertise} forces the team to rely mainly on their
personal abilities, even though asking opinions to mentors is reported to be a
viable practice to maintain feasible objectives. Furthermore \textit{teams work
under constant pressure}, \textit{``up to 16 hours per day''}, %MUN: add Cx to
%this statement since it is a citation?  
%CG: not fixed. It represents a subcategory, expressed by the 70% of the  
%interviewed startups 
%MUN1: and how should the reader know that? See my problem with italics: you  
%use italics for both interview statements and (sub-)categories. 
%CG: skip as acked 
mainly constrained by a tight \textit{time shortage}.

Finally, startups present founders-centric structures, and especially in the
early-stage, it is clear that \textit{CTO/CEO background has high-impact} on the
company's development approach. For instance, in case of an academic background,
the CTO encourages the introduction of some architectural design before the
development phase. Moreover, even though the CTO/CEO starts guiding initially
the development process, most of the decision are taken consensually by all
members of the team. Then, the CTO/CEO decides only in particular situations
where some conflicts occur.
%------------------------------------------------------------------------------%
\subsection{Evolutionary approach}\label{res:gsm:cat2} 
Among different approaches to development, startups prefer to build an initial  
prototype and iteratively refine it over time, similarly to the concept of
\textit{``evolutionary prototyping''} \cite{EvProt}. The reason of using this
approach is that startups want to validate the product against the market as
soon as possible, finding the proper product/market fit. Indeed, they can focus
on developing only parts of the system they want to validate instead of working
on developing a whole new system. Then, as the prototype is released, users
detect opportunities for new functionalities and improvements, providing their
feedback to developers.

@@

The objective of this evolutionary approach is to avoid wasting time in  ``over-
engineering the system'' and building complex functionalities that have  not
been tested on real users. By releasing a very small number of good-enough
functionalities (see CAT3) the startup can verify the suitability of the
features and understand how to adjust the direction of product development
towards actual users' needs. The first version of the product is typically a
prototype which contains the basic functionalities developed with the least
possible effort that can validate critical features, enabling the startup's
survival in the short term (recalling  Ries' definition of \textit{Minimum
Viable Product} (MVP) \cite{Ries2011}).

@@

%------------------------------------------------------------------------------%
\subsection{Product quality has low priority}\label{res:gsm:cat3}
The interests of software startups, related to the product, are mainly  
concentrated on building a  \textit{limited number of suitable functionalities} 
rather than fulfilling restrictive non-functional requirements. This strategy  
allows them to quickly release simple products with less need of preliminary  
architectural studies.

@@

Finally, startups implement a \textit{limited number of suitable
functionalities}, which in view of \textit{limited human resources} and
\textit{time shortage}, represent a viable strategy to focus on shortening
time-to-market.
%------------------------------------------------------------------------------%
\subsection{Speed-up development}\label{res:gsm:cat1} 
As remarked initially, \textit{speed-up development} represents the core  
category of the GSM. Firmly grounded as the primary objective of startups, it  
shows the most important characteristic of developing software in the early  
stage.

@@

Finally the constant pressure under which the company regularly operates, leads
the team to often \textit{work overtime to meet deadlines}. But as reported by
practitioners, such way of working can be an effective strategy only in the
short term since it can lead to poorly maintainable code and developer burnouts
in the long run.
%------------------------------------------------------------------------------%
\subsection{Accumulated technical debt}\label{res:gsm:cat5} 
Startups achieve a very high development speed by radically ignoring aspects  
related to documentation, structures and processes. However, although these  
aspects are not considered important in the very first stages, they will become
increasingly more important for the productivity in the long-term, as we
illustrate in CAT6.

@@

Another important factor which contributes to \textit{accumulate the technical
debt} is the general \textit{lack of architectural design}, substituted by
\textit{high-level mockups and low-precision diagrams}, \textit{using
modularization and frameworks from day 1} and \textit{describing critical
interactions with third-party components only}. In particular, the use of  well-
known standards, frameworks and conventions limits the need of formal UML
\cite{UML} diagrams and documentation, and provide a minimum level of
maintainability costs. Additionally, since quality aspects are not a main
concern (see \textit{limited number of functionalities} and \textit{efficiency
emerges after using the product} in CAT3) having a well-structured architecture
remains a secondary priority, and design is conducted only when strictly
necessary.

@@

Furthermore, a rigid project management is perceived as \textit{``waste of
time''} which hinders the development speed since the \textit{uncertainty makes
formal scheduling pointless} (C9  reports that \textit{``initial chaos helps to
develop faster''}). Startups' \textit{minimal project management}  is supported
by keeping: \textit{internal milestones short and informal},  \textit{low-
precision task assignment mechanism}  and a very low attention for  project
metrics (paraphrasing C13, \textit{``the only track of progress was made  by
looking at closed tickets''}). In this context \textit{only a final release
milestone is viable}, which helps practitioners to remain focused on short term
goals and put new features in production. To further \textit{speed-up
development}, startups \textit{simplify issue management by integrating it with
feature management}. Then, the \textit{limited number of functionalities} and
the {use of standard/known technologies} with a \textit{simple workflow} are the
main reasons for  not establishing a heavy project management process since the
project and  technologies can be autonomously managed by the development team
(with  characteristics of: co-location, self-organization and very small team
size).  %MUN: do you mean small team size? %CG: fixed

Finally, one of the categories, which most contributes in growing the
\textit{accumulated technical debt}, is the substantial  use of informal and
verbal communication channels on a daily basis. The high co-location and the
fast paced development approach increase the volume of \textit{tacit knowledge}
and severe lack of any kind of documentation.  C4 observes in this regard that:
\textit{``[\ldots] the issue of having documentation and diagrams out of the
source code is that you need to update them every time you change something.
There is no time for it. Instead, there is a huge pay off in having a code that
is understandable itself.''} This approach is supported by the fact that
startups make use of   \textit{simple and informal workflow},
\textit{standard/known technologies},  \textit{very small and co-located
development team} with \textit{limited need of  formalities}.
%------------------------------------------------------------------------------%
\subsection{Initial growth hinders performance}\label{res:gsm:cat6} 
The lack of attention given in the first phases to engineering activities  
allows startups to ship code extremely quickly (see CAT5). However, if  
successful, the initial product becomes more and more complex over time, the  
number of users increases and the company size starts to grow. Under these  
circumstances the need of controlling the initial chaos forces the development  
team to return the \textit{accumulated technical debt} instead of focusing on  
new users' requests. Hence, \textit{initial growth hinders performance} in terms 
 of number of new user stories implemented in a certain amount of time, which  
brings new suitable functionalities to the users.

@@

Therefore, startups begin to \textit{partially replace informal communication
with traceable systems} and \textit{introduce basic metrics for measuring
project and team progress} in order to establish an initial structured workflow.
In conclusion, because the increasing number of users causes issues in
scalability and reliability, the \textit{need of more structured workflow}
becomes of major relevance. C11 states: \textit{``Yet, it is still better to
have a reasonable drop-down in performance when team grows than lose time in the
beginning''}.
%------------------------------------------------------------------------------%
\subsection{Theory}\label{res:gsm:th}
In order to explain and understand the development strategies in early-stage  
software startups we construct the theory generated and supported by the above  
presented GSM:

@@

In this section we discuss the validity of the GSM by means of 
cross-methodological observations, as discussed in section \ref{rm:val}. 
%MUN: expand this discussion a bit (motivation why you do this like you do it)  
%and move to Section 3.5. Then write here just that you perform theory  
%validation as discussed in Section 3.5 (and remove the bullet list below). 
%CG:fixed 
%MUN1: Ok, but fix section 3.5... you see that you don't motivate 5.1-5.4 at  
%all there and that the current content of 3.5 is not related to what you  
%actually do here as validation? CG: skip as acked

@@

Starting from the Coleman's work, we refer to him since only he has,  to the
best of our knowledge, conducted a similar research in startups' context, even
though with a different focus. Coleman investigated factors in software
development that hinder initiatives of software process improvement (SPI) in a
later stage, representing also companies in expansion phase with more than 100
employees. %MUN: so he did not do research  in a startup context? You need to
%clarify what you mean by "more mature companies"... I assume you still mean
%startups, but startups at a later stage, i.e. after first product release.
%CG: fixed

@@

As also reported in the GSM, the definition of a \textit{``minimum process''}
is not a matter of poor knowledge and training, but rather it is the necessity
of operating with solutions that let the company move faster.  \textit{``One-
size-fits-all''} solutions have always found difficulty in  penetrating small
software organizations \cite{Staples2007}. When startups began  establishing any
SPI process, they experienced \textit{process erosions}   \cite{Coleman2008}
%MUN:  %add ref (Coleman?). You should add, for each category you use from
%Coleman, add the reference to the paper. This makes it clear that the category
%comes from him (and is not one of our categories) %CG: fixed 
, which resolved to a barely sufficient workflow to satisfy the organizational  
business needs.

Then, software startups favor the use of agile principles in support of
\textit{creativity} and \textit{flexibility} instead of SPI methods, whereby
processes need to be predictable and repeatable. Nimble and ad-hoc solutions
prevent the use of heavy bureaucracy and formal communication strategies %MUN:
%seems to contradict each other: "ad-hoc solutions prevent the use of heavy
%bureaucracy AND fast communication strategies" I agree that ad-hoc solutions
%do not produce heavy bureaucracy, but they allow fast communication
%strategies... I may be wrong, but then you need to explain it better %CG:fixed,
%you are right :) ...   
, even though the accumulated tacit knowledge is hard to manage and to transfer 
to new hires as also discussed in category \textit{CAT6}.

@@

\subsubsection{Market requirements and application type}  
Market requirements \cite{Coleman2007} and application type \cite{Sutton2000,  
Coleman2008,Coleman2008a} are two further potential confounding factors to the  
GSM. Their main impact is related to the adoption of flexible and reactive  
solutions for the development process. In particular, they refer to the  
necessity of fulfillment of quality concerns that goes beyond scalability and UX 
defined by the GSM. Especially when requirements are rigidly imposed or the  
application domain is well-known, providing low-quality products to final users 
might determine the failure of a startup.
